Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-11-01 latex/paper.tex
Final version

\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
\IEEEauthorblockA{
Delft University of Technology\\
Delft, The Netherlands\\
Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
}
}
\maketitle

@@

Due to a variety of reasons, including the advent of cloud computing, the
rising rate of information production and the necessity to reach the market
fast, currently, large corporations and start-ups alike are investigating
alternative programming and information storage models. As a result, during
the last few years, the practical software engineering field is witnessing a
noticeable shift towards functional programming. Scripting languages, notably
Javascript and Ruby, pioneered the introduction of functional concepts, such as
closures and lambda functions, to mainstream programming. A new wave of
programming languages, developed to overcome the expressiveness and
complexity limitations exhibited in mainstream languages, have promoted
functional constructs, such as type safe pattern matching, higher order
functions and single assignment variables, to first class citizens (Scala).
New, purely functional, languages have emerged to fill in the remaining gaps
(F\#, Clojure), often introducing significant advancements in their field of
specialisation (such as Erlang in distributed fault-tolerant systems).
Finally, large scale information processing systems such as Map-Reduce and
domain specific languages such as LINQ have integrated functional concepts to
ease the expression of computations.

@@

The functional programming course is elective at the MSc curriculum. Students
from all departments in the Electrical Engineering, Mathematics and Computer
science faculty are allowed to participate. For practical reasons, the
enrollment to the course was limited to 15 participants. This permitted
us to get to know the students individually and to provide them with
personalized support.

From the students that enrolled, most have received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce~\cite{Dean04} in the web information systems
course). Two of the students were majoring in computer engineering, which meant
that their programming experience was restricted. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

The expected diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. For this reason, we decided
to adopt a hands-on-first approach; the students would have to learn by
flexing their programming muscles, instead of being gently introduced to 
the theory through toy examples.

@@

Functional programming is arguably the oldest programming paradigm. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus~\cite{Baren84}). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theoretical background, understanding
it usually leads to more elegant algorithmic solutions. Functional
programming teaching is also associated with advanced type systems; the flagship
functional programming languages (Haskell and the {\sc ml} family) both offer
very sophisticated support for types. Consequently, teaching the full spectrum
of functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional techniques.

@@

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
tough process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important student feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, {\sc f\#} and Clojure are functional languages working
on familiar development platforms. Apart from staying compatible with the
existing literature, there is no practical reason that necessitates teaching of
functional programming strictly in a language like Haskell or {\sc ml}.

@@

The aim of the course was to teach the principles of pure functional
programming, and the corresponding Category theoretical principles. 
More specifically, the educational purposes of the course were:

@@

  \item To introduce students to basic functional programming concepts, such
    as higher order functions, monads and advanced type systems.

  \item To introduce students to the approach of viewing data processing
    problems as a series of function applications.

@@

  \item To

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, and laboratory sessions. In total, the
course consisted of 14 hours of lectures and 8 hours of lab work.
The students
also had to carry out a series of intermediate assignments and a final project.
The students would be evaluated by their performance on the final assignment. 
The lectures took place in a quarter (half semester), during 3 intense 
weeks; the students had another 6 weeks to finish their final projects.

@@

The assignments were open ended, as there was no fixed set of targets for
each one of them. The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to present a working
demo by the end of the course time.

@@

The purpose of the lab sessions where to help students carry out their final
assignments and explain better concepts that might not be clear to students. 
Students convened every week, presented their progress and discussed with 
each other and with the lab demonstrator design and practical issues that
arised in the implementation of their projects. To further motivate students
to work on their projects, one of the lab sessions was converted to a full
day coding sprint.

@@

The focus of teaching was to present functional programming from a practical
aspect; we were mostly interested to teach students what they can do rather than
the theory behind functional constructs. An important distinction
that was made early on was that the every day world is imperative; therefore
while functional programming can be a great tool for thinking about a problem, a
pure functional solution cannot reflect on the real world. As a consequence it
is usually best to mix programming paradigms, i.e. imperative or object-oriented
programming for state representation and functional programming for data
processing. In our experience, this distinction helped students understand
that they are already doing functional programming without noticing;
enforcing data immutability to handle multi-threading problems or passing
function arguments in scripting languages are applications of functional
programming principles. The moment students realized this fact, their
attention was captured to the lectures.

During both lectures and labs, all non-whiteboard examples where actually
entered in a real programming environment (LinqPad or the Scala command-line
read-evaluate loop) and the results of the evaluation where discussed with the
students. Participation was encouraged by modifying the examples and asking the
students to perform the evaluation, before the example was executed. The
examples varied from list manipulation with higher-order functions to 
a toy actor model example.

@@

\begin{figure*}
\centering
\label{fig:teapots}
\subfigure[Input drawing, rendered with Java graphics]{
\includegraphics[scale=0.173]{teapot-reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposition in Scala, rendered with Java graphics]{
\includegraphics[scale=0.17]{teapot-java-canvas.png}
\label{fig:teapot-scala}
}
\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
elements]{
\includegraphics[scale=0.2]{teapot-html-divs.png}
\label{fig:teapot-html-divs}
}
\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
\includegraphics[scale=0.31]{teapot-html-canvas.png}
\label{fig:teapot-html-canvas}
}
\caption[]{Example results from the teapot exercise.}
\end{figure*}

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles. The students
had to come up with the decomposition method (using some form of analytic
geometry), a decomposition restriction criterion to stop the decomposition when
triangles could not be rendered on screen and a method to recursively apply the
above mentioned transformations on the input data. As always, the students could
decide the implementation language of their choice. The exercise was given as a
mid-week homework assignment, between the Monday and the Friday lectures.

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of a pure functional
language. In spite of the advice, students were eager to test-drive the newly
acquired knowledge in a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

@@

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. The lecture on Map/Reduce featured an invited
talk by Ralf L\"ammel, who presented his formulation of Map/Reduce using
functional programming principles~\cite{Lamme08} and explained how those can be
applied to process semi-structured data. The talk resonated to students; in
their projects, students used implicit, through the selected language's library
collection methods, or explicit, by formalizing data processing as a series of
in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
Out of personal interest, one student even implemented L\"ammel Map/Reduce
formulation in Scala, and through the language's extension mechanisms, adapted
it for use as an extension method by any Scala collection type, including
parallel collections.

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively in a tight deadline to deliver them. To ensure
the timely delivery of the student projects, we organized an one day
coding sprint, during the students where request to i) set an explicit target 
ii) write code for 7 hours with a short break iii) present a 5 minute 
demo of their work at the end of the day. All project teams where expected to
work in the same room, while communication among the teams was facilitated
by the lab demonstrator, to avoid unnecessary distractions.

The sprint helped students kick start their projects.

\subsection{The student projects}

As expected, the performance of students in their projects varied.

\section{Conclusion}

Teaching the concepts of solving problems algorithmically can be a daunting
task~\cite{Futsc06}; especially more so, when the subjects are already in a
specific mindset. The assumption of the presented course was that imperative and
functional programming are really the two sides of the same coin; by focusing on
how students can apply rather than just be taught concepts of functional
programming using the, typically imperative, languages they already know, allows
them to better appreciate the strong points and weaknesses of each paradigm. In
our experience, the approach has been successful. Both throughout the initial
homework assignments and in their projects, the students demonstrated a high
degree of appreciation of functional programming concepts. Above all, we
believe that it was the hands-on approach through tough homework exercises
employed in this course that allowed students to sharpen their skills and
understand the concepts that were taught during the course.

\section*{Acknowledgements and Availability}

We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}.

>>>>

\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
\IEEEauthorblockA{
Software Engineering Research Group\\
Delft University of Technology\\
Delft, The Netherlands\\
Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
}
}
\maketitle

@@

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating alternative
programming and information storage models. As a result, during the last few
years, the practical software engineering field is witnessing a noticeable shift
towards functional programming. Scripting languages, notably Javascript and
Ruby, pioneered the introduction of functional concepts, such as closures and
lambda functions, to mainstream programming. A new wave of programming
languages, developed to overcome the expressiveness and complexity limitations
exhibited in mainstream languages, have promoted functional constructs, such as
type safe pattern matching, higher order functions and single assignment
variables, to first class citizens (Scala).  New, purely functional, languages
have emerged to fill in the remaining gaps (F\#, Clojure), often introducing
significant advancements in their field of specialisation (such as Erlang in
distributed fault-tolerant systems).  Finally, large scale information
processing systems such as Map/Reduce~\cite{Dean04} and domain specific
languages such as {\sc linq} have integrated functional concepts to ease the
expression of computations.

@@

The functional programming course is elective at the MSc level. Students
from all departments in the Electrical Engineering, Mathematics and Computer
science faculty are allowed to participate. For practical reasons, the
enrollment to the course was limited to 15 participants. This permitted
us to get to know the students individually and to provide them with
personalized support.

From the students that enrolled, most have received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce in the web information systems
course). Two of the students were majoring in computer engineering, which meant
that their programming experience was restricted. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

The diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. For this reason, we decided
to adopt a hands-on-first approach; the students would have to learn by
flexing their programming muscles, instead of being introduced to 
the theory through toy examples.

@@

Functional programming is arguably the oldest programming paradigm. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus~\cite{Baren84}). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theory, understanding it usually leads
to more elegant algorithmic solutions. Functional programming teaching is also
associated with advanced type systems; indeed, the flagship functional
programming languages (Haskell and the {\sc ml} family) both offer very
sophisticated support for types. Consequently, teaching the full spectrum of
functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional
techniques.

@@

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
thought process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important student feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, {\sc f\#} and Clojure are functional languages
implemented on familiar development platforms. Apart from staying compatible
with the existing literature, there is no practical reason that necessitates
teaching of functional programming strictly in a language like Haskell or {\sc
ml}.

@@

The high level goal of the course was to teach the principles of functional
programming, and the corresponding Category theoretical principles. 
More specifically, the educational purposes of the course were:

@@

  \item To introduce students to basic functional programming concepts,
    higher order functions, monads and advanced type systems.

  \item To introduce students to the approach of expressing data processing
    problems as a series of function applications.

@@

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, and laboratory sessions. In total, the
course consisted of 14 hours of lectures and 8 hours of lab work.
The students
also had to carry out a series of homework assignments and a final project.
The students would be evaluated by their performance on the final assignment. 
The lectures took place in a quarter (half semester), during 3 intense 
weeks; the students had another 6 weeks to finish their final projects.

@@

To counterbalance the differences in student experience level, 
the assignments were open ended; students could drive their projects as
far as they where willing to.
The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to present a working
demo by the end of the course time.

@@

The purpose of the lab sessions where to help students carry out their final
assignments and further explain concepts that might not be clear to students. 
Students convened every week, presented their progress and discussed with 
each other and with the lab demonstrator design and practical issues that
arised in the implementation of their projects. To further motivate students
to work on their projects, one of the lab sessions was converted to a full
day coding sprint.

@@

The focus of teaching was to present functional programming from a practical
aspect; we were mostly interested to teach students what they can do rather than
the theory behind functional constructs. An important distinction that was made
early on was that the world is imperative; therefore while functional
programming can be a great tool for thinking about a problem, a pure functional
solutions cannot reflect on the real world. As a consequence, it is usually best
to mix programming paradigms, i.e. imperative or object-oriented programming for
state representation and functional programming for data processing. In our
experience, this distinction helped students understand that they are already
doing functional programming without noticing; enforcing data immutability to
handle multi-threading problems or passing function arguments in scripting
languages are applications of functional programming principles. The moment
students realized this fact, their attention was captured to the lectures.

During both lectures and labs, all non-whiteboard examples where actually
entered in a real programming environment (LinqPad or the Scala command line
read-evaluate loop) and the results of the evaluation where discussed with the
students. Participation was encouraged by modifying the examples and asking the
students to perform the evaluation, before the example was executed. The
examples varied from list manipulation with higher-order functions to 
a toy actor model example.

@@

\begin{figure*}
\centering
\label{fig:teapots}
\subfigure[Input drawing, rendered with Java graphics]{
\includegraphics[scale=0.173]{teapot-reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposition in Scala, rendered with Java graphics]{
\includegraphics[scale=0.17]{teapot-java-canvas.png}
\label{fig:teapot-scala}
}
\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
elements]{
\includegraphics[scale=0.195]{teapot-html-divs.png}
\label{fig:teapot-html-divs}
}
\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
\includegraphics[scale=0.295]{teapot-html-canvas.png}
\label{fig:teapot-html-canvas}
}
\caption[]{Example results from the teapot exercise.}

\vspace{-1.5em}
\end{figure*}

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles. The students
had to come up with the decomposition method (using an analytic geometry
method), a decomposition termination criterion to stop the decomposition when
triangles are too small to be rendered on screen and a method to recursively
apply the above mentioned transformations on the input data. As always, the
students could decide the implementation language of their choice. The exercise
was given as a mid-week homework assignment, between the Monday and the Friday
lectures.

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of learning a new 
language. In spite of the advice, students were eager to test-drive the newly
acquired knowledge using a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

@@

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. The lecture on Map/Reduce featured an invited
talk by Prof. Ralf L\"ammel, who presented his formulation of Map/Reduce using
functional programming principles~\cite{Lamme08} and explained how those can be
applied to process semi-structured data. The talk resonated to students; in
their projects, students used implicit, through the selected language's library
collection methods, or explicit, by formalizing data processing as a series of
in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
Out of personal interest, one student even implemented L\"ammel's Map/Reduce
formulation in Scala, and through the language's extension mechanisms, adapted
it for use as an extension method by any Scala collection type, including
parallel collections.

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively in a tight deadline to deliver them. To ensure the
timely delivery of the student projects, we organized an one day coding sprint,
during which the students were requested to i) set explicit targets ii) write
code for 7 hours with a short break iii) present a 5 minute demo of their work
at the end of the day. All project teams where expected to work in the same
room, while communication among the teams was facilitated by the lab
demonstrator, to avoid unnecessary distractions.

Interestingly, the coding sprint was organized on student request. The reason
was that pressure from other courses did not allow them to be together as teams.
The sprint taught students the importance of setting realistic development goals
and the necessity of iterative development. Even though all teams were off
target at the end of the day, most were able to create a rough prototype in less
than 7 hours, using functional programming techniques.

%\subsection{The student projects}
%
%For implementing their projects, the students selected languages and frameworks
%that are currently emerging and popular among non-traditional software
%development environments. All students did their projects in either Scala or
%Javascript, with one exception, using frameworks such as Akka (for actors),
%Scalatra (for implementing {\sc rest} {\sc api}s) or Node.js (for server-side
%processing). From the implementations it appears that students did apply
%their 
%
\section{Conclusion}

Teaching the concepts of solving problems algorithmically can be a daunting
task~\cite{Futsc06}; especially more so, when the subjects are already in a
specific mindset. The assumption of the presented course was that imperative and
functional programming are really the two sides of the same coin; by focusing on
how students can apply rather than just be taught concepts of functional
programming using the, typically imperative, languages they already know, allows
them to better appreciate the strong points and weaknesses of each paradigm. In
our experience, the approach has been successful. Both throughout the initial
homework assignments and in their projects, the students demonstrated a high
degree of appreciation of functional programming concepts. Above all, we
believe that it was the hands-on approach 
employed in this course that allowed students to sharpen their skills and
understand the concepts that were taught during it.

\section*{Acknowledgements and Availability}
We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}. This work is partially supported by
Marie Curie {\sc ief} 298930 -- {\sc sefunc}.
